Fuels, fire weather, and Rx fire behavior in North Dakota

### 2021-02-18 writeups/FuelsWeatherFirebehavior2.Rmd
robust imputation update

---
title: Fuels, fire weather, and Rx fire behavior in North Dakota
subtitle: 
date: '`r Sys.Date()`'
author: DAM
output:
  html_document:
    keep_md: false
    toc: true
    toc_float: true
    theme: united
    highlight: tango
---

```{r setup, echo=FALSE, warning=FALSE, message = FALSE, results='hide'}
knitr::opts_chunk$set(message = FALSE, warning=FALSE, 
               echo=FALSE, eval=TRUE, fig.path='figures/', dev=c('png', 'pdf')) 
FilePath = "C:/Users/devan/GitHubProjects/SpatialFireBehavior"
AnalysisData <- readr::read_csv(paste0(FilePath, "/data/AnalysisDataKY.csv"))
load(paste0(FilePath, '/r objects/response_CIs.Rdata'))
load(paste0(FilePath, '/r objects/imp_raw.Rdata'))
load(paste0(FilePath, '/r objects/pca_ev.Rdata'))
source(paste0(FilePath, '/r script/CustomThemes.R'))
source("https://raw.githubusercontent.com/devanmcg/IntroRangeR/master/11_IntroMultivariate/ordinationsGGplot.R")
```

```{r}
# Packages
  pacman::p_load(tidyverse, mice, lubridate, gridExtra, vegan)
```

@@

```{r data_prep, eval=FALSE}
###
# Note: Data prep steps included for reference. 
# Data used in analysis, AnalysisDataKY.csv, has been modified outside of
# this script to update missing data from another spreadsheet. 
 AllData <-  
    read_csv(paste0(FilePath, "/data/fromMZ/CompiledData2.csv")) %>%
    filter(location != "OAK") %>%
      mutate(date = as.Date(date, format = "%m/%d/%Y"),
               L = str_remove(location, "REC"), 
               B = str_sub(block, 1,3), 
               Ps = str_replace(pasture, "[.]", ""), 
               Ps = str_sub(Ps, 1,2), 
               patch = str_replace(patch, "[.]", ""),
               y = format(date, "%y")) %>%
        unite("FireCode", c(L,B,Ps,patch,y), sep=".") %>%
    mutate(time = str_remove(MaxTempTime, "[.]+[0-9]"))%>%
    unite(timestamp, c(date, time), sep = " ") %>%
    mutate(timestamp = as.POSIXct(timestamp, format = "%Y-%m-%d %H:%M:%S")) %>%
        select(FireCode, timestamp, plot, array, TC, MaxC, 
               AirTemp, RH, dpC, MaxWindSpeed, 
               LAI, FMC, KgHa)

@@

```{r fig.width = 10, fig.height = 8, fig.cap="A scatterplot matrix of all raw data used in this analysis."}
# Scatterplot matrix  
AnalysisData %>%
  select(-LAI) %>%
  GGally::ggpairs(columns = 9:18, 
                  upper = 'blank')
```

```{r dist_gg, fig.width = 8, fig.height =8, fig.cap = "Gamma distributions are fit to all response variables. Two outliers above 600 deg C were removed from analysis on soil maximum temperature."}
# Distributions of response variables
  ros_dist <-
  AnalysisData %>% 
    ggplot(aes(x=ros)) + theme_bw(14) + 
    geom_histogram(aes(y=..density..),      
                   binwidth=0.5,
                   colour="black", 
                   fill="lightgreen") +
    geom_density(alpha=0.2, 
                 fill="#FF6666") 
  tempC_dist <- 
  AnalysisData %>% 
    ggplot(aes(x=MaxC)) + theme_bw(14) + 
    geom_histogram(aes(y=..density..),      
                   binwidth=10,
                   colour="black", 
                   fill="lightgreen") +
    geom_density(alpha=0.2, 
                 fill="#FF6666") 
  soilC_dist <- 
  AnalysisData %>% 
  ggplot(aes(x=SoilMaxC)) + theme_bw(14) + 
  geom_histogram(aes(y=..density..),      
                 binwidth=10,
                 colour="black", 
                 fill="lightgreen") +
  geom_density(alpha=0.2, 
               fill="#FF6666") 
  gridExtra::grid.arrange(tempC_dist, soilC_dist, 
                          ros_dist, ncol=2)
```

@@

```{r missing_table}
# MICE operations 
#
# Pattern of missing data
AnalysisData %>%
  select(-FireCode, -location, -block, -pasture,-patch,
         -year, -plot, -array) %>%
    mice::md.pattern(plot = F) %>%
  pander::pander(caption="Table of missing values by variable.") 
```

@@

```{r imp, eval=FALSE}
# Calculate imputed datasets
  imp <- AnalysisData %>% 
          select(-LAI) %>%
              mutate(across(plot:array, ~as.character(.))) %>%
               mice(m=500, seed = 23109, print=F)

# Return completed datset   
  imp_raw <- 
    complete(imp) %>%
            as_tibble() %>% 
              mutate(across(location:array, as.factor))

# Scale response variables for regression analysis   
  imp_sc <- 
    imp_raw %>%
     mutate_at(vars(AirTemp:tHa), ~as.numeric(scale(., center=F)))
```

@@

Mixed-effect regression models fit with `glmmADMB::glmmadmb` and constructed as follows:

@@

```{r ADMB_mods, eval=FALSE}
# Max soil surface temperature 
  # Fit model
    soil_ADMB <- glmmADMB::glmmadmb(SoilMaxC ~ RH + MaxWindSpeed +
                            FuelMoisture + tHa,
                          random = ~ 1|location/year/plot ,
                          data=filter(imp_sc, SoilMaxC <= 3.9),
                          family="gamma",
                          link="log",  
                          admb.opts=glmmADMB::admbControl(shess = TRUE, 
                                                          noinit = TRUE, 
                                                          impSamp = 10) )
  # Retrieve regression results
    soil_terms <-   
      confint(soil_ADMB) %>%
      as.data.frame %>%
      rownames_to_column("term") %>%
      rename(lwr = `2.5 %`, upr = `97.5 %`) %>%
      full_join(
        coef(soil_ADMB) %>%
          as.data.frame() %>% 
          rownames_to_column("term") %>%
          rename(estimate = '.') )

# Maximum aboveground temp (plant canopy)  
  # Fit model
    canopy_ADMB <- glmmADMB::glmmadmb(MaxC ~ RH + MaxWindSpeed +
                                        FuelMoisture + tHa,
                          random = ~ 1|location/year/plot ,
                          data=imp_sc,
                          family="gamma",
                          link="log",  
                          admb.opts=glmmADMB::admbControl(shess = TRUE, 
                                                          noinit = TRUE, 
                                                          impSamp = 10))
  # Retrieve regression results
  canopy_terms <-   
    confint(canopy_ADMB) %>%
    as.data.frame %>%
    rownames_to_column("term") %>%
    rename(lwr = `2.5 %`, upr = `97.5 %`) %>%
    full_join(
      coef(canopy_ADMB) %>%
        as.data.frame() %>% 
        rownames_to_column("term") %>%
        rename(estimate = '.') )

# Rate of spread 
  # Fit model 
    ros_ADMB <- glmmADMB::glmmadmb(ros ~ RH + MaxWindSpeed +
                                    FuelMoisture + tHa,
                          random = ~ 1|location/year/plot ,
                          data=imp_sc,
                          family="gamma",
                          link="log",  
                          admb.opts=glmmADMB::admbControl(shess = TRUE, 
                                                          noinit = TRUE, 
                                                          impSamp = 10))
  # Retrieve regression results  
    ros_terms <-   
      confint(ros_ADMB) %>%
      as.data.frame %>%
      rownames_to_column("term") %>%
      rename(lwr = `2.5 %`, upr = `97.5 %`) %>%
      full_join(
        coef(ros_ADMB) %>%
          as.data.frame() %>% 
          rownames_to_column("term") %>%
          rename(estimate = '.') )

# Compile results

    response_CIs <-   
      bind_rows(
        mutate(soil_terms, response = 'Temp. at surface'), 
        mutate(canopy_terms, response = 'Temp. above surface'),
        mutate(ros_terms, response = "Rate of spread") 
      ) %>%
      as_tibble()
```

@@

**Rate of spread** significantly increased with windspeed ($z = 3.24, P < 0.01$) and decreased with greater fuel moisture content ($z = -5.84, P < 0.001$). 
**Aboveground temperatures** in the plant canopy were lower with greater fuel moisture content ($z = -6.45, P < 0.001$) and increased with fuel load ($z = 5.49, P < 0.001$) and wind speed ($z = 2.82, P < 0.01$).
**Temperature at the soil surface** had a statistically-significant inverse relationship with relative humidity ($z = -2.52, P = 0.01$) and fuel moisture ($z = -3.22, P < 0.01$), which likely indicates higher-intensity fires that burn down through litter.

```{r CI_gg, fig.width=8, fig.height = 3.75}
# Plot confidence intervals
response_CIs %>%
  filter(term != '(Intercept)') %>%
  as_tibble() %>%
  mutate(term = recode(term, tHa = 'Fuel load', 
                             MaxWindSpeed = 'Wind speed', 
                             RH = 'Relative humidity', 
                             FuelMoisture = 'Fuel moisture'))  %>%
  ggplot(aes(x = reorder(term, abs(estimate), max))) + theme_bw(16) + 
    geom_hline(yintercept = 0, color="black", linetype = 2) +
    geom_errorbar(aes(ymin = lwr, ymax = upr), 
                  size = 1.1, width = 0.2, 
                  color = "blue") +
    geom_point(aes(y = estimate), pch = 21, 
               stroke = 1.5, size = 4,
               color = "blue", fill="lightblue") +
    labs(x = '', 
         y = 'Regression coefficient with 95% CI') + 
    coord_flip(ylim=c(-1.5,2.4)) + 
    facet_wrap(~ response) + 
    theme(axis.text.y = element_text(color="black", size = 14), 
          strip.text = element_text(size = 14), 
          panel.grid.major.y = element_blank())
```

>>>>

---
title: Fuels, fire weather, and Rx fire behavior in North Dakota
subtitle: Robust imputation with `mice`
date: '`r Sys.Date()`'
author: DAM
output:
  html_document:
    keep_md: false
    toc: true
    toc_float: true
    theme: united
    highlight: tango
---

```{r setup, echo=FALSE, warning=FALSE, message = FALSE, results='hide'}
knitr::opts_chunk$set(message = FALSE, warning=FALSE, 
               echo=FALSE, eval=TRUE, fig.path='figures/', dev=c('png', 'pdf')) 
FilePath = "C:/Users/devan/GitHubProjects/SpatialFireBehavior"
AnalysisData <- readr::read_csv(paste0(FilePath, "/data/AnalysisDataKY.csv"))
load(paste0(FilePath, '/r objects/response_CIs.Rdata'))
load(paste0(FilePath, '/r objects/imp_raw.Rdata'))
load(paste0(FilePath, '/r objects/pca_ev.Rdata'))
source(paste0(FilePath, '/r script/CustomThemes.R'))
source("https://raw.githubusercontent.com/devanmcg/IntroRangeR/master/11_IntroMultivariate/ordinationsGGplot.R")
# Load for markdown 
pacman::p_load(tidyverse, mice, lubridate, gridExtra, vegan)

response_CIs <- response_CIs %>%
  filter(term != '(Intercept)') %>%
  as_tibble() %>%
  mutate(term = recode(term, tHa = 'Fuel load', 
                             MaxWindSpeed = 'Wind speed', 
                             RH = 'Relative humidity', 
                             FuelMoisture = 'Fuel moisture')) 
```

```{r, eval = FALSE}
# Packages & script for analysis
  pacman::p_load(tidyverse, mice, broom.mixed, lubridate, gridExtra, vegan)
  source('C:/Users/devan/GitHubProjects/SpatialFireBehavior/r script/mipo.R')
```

@@

Figures generated by the script below and included in this document are [available on GitHub](https://github.com/devanmcg/SpatialFireBehavior/tree/main/writeups/figures), as well.

```{r data_prep, eval=FALSE}
###
# Note: Data prep steps included for reference. 
# Data used in analysis, AnalysisDataKY.csv, has been modified outside of
# this script to update missing data from another spreadsheet. 
 AllData <-  
    read_csv(paste0(FilePath, "/data/fromMZ/CompiledData2.csv")) %>%
    filter(location != "OAK") %>%
      mutate(date = as.Date(date, format = "%m/%d/%Y"),
               L = str_remove(location, "REC"), 
               B = str_sub(block, 1,3), 
               Ps = str_replace(pasture, "[.]", ""), 
               Ps = str_sub(Ps, 1,2), 
               patch = str_replace(patch, "[.]", ""),
               y = format(date, "%y")) %>%
        unite("FireCode", c(L,B,Ps,patch,y), sep=".") %>%
    mutate(time = str_remove(MaxTempTime, "[.]+[0-9]"))%>%
    unite(timestamp, c(date, time), sep = " ") %>%
    mutate(timestamp = as.POSIXct(timestamp, format = "%Y-%m-%d %H:%M:%S")) %>%
        select(FireCode, timestamp, plot, array, TC, MaxC, 
               AirTemp, RH, dpC, WindSpeed, 
               LAI, FMC, KgHa)

@@

```{r fig.width = 10, fig.height = 8, fig.cap="A scatterplot matrix of all raw data used in this analysis."}
# Scatterplot matrix  
AnalysisData %>%
  select(-LAI, -JD) %>%
  GGally::ggpairs(columns = 9:18, 
                  upper = 'blank')
```

```{r dist_gg, fig.width = 8, fig.height =8, fig.cap = "Gamma distributions are fit to all response variables. An outlier well above 600 deg C were removed from analysis on soil maximum temperature."}
# Distributions of response variables
  ros_dist <-
  AnalysisData %>% 
    ggplot(aes(x=ros)) + theme_bw(14) + 
    geom_histogram(aes(y=..density..),      
                   binwidth=0.5,
                   colour="black", 
                   fill="lightgreen") +
    geom_density(alpha=0.2, 
                 fill="#FF6666") 
  tempC_dist <- 
  AnalysisData %>% 
    ggplot(aes(x=MaxC)) + theme_bw(14) + 
    geom_histogram(aes(y=..density..),      
                   binwidth=10,
                   colour="black", 
                   fill="lightgreen") +
    geom_density(alpha=0.2, 
                 fill="#FF6666") 
  soilC_dist <- 
  AnalysisData %>% 
  ggplot(aes(x=SoilMaxC)) + theme_bw(14) + 
  geom_histogram(aes(y=..density..),      
                 binwidth=10,
                 colour="black", 
                 fill="lightgreen") +
  geom_density(alpha=0.2, 
               fill="#FF6666") 
  gridExtra::grid.arrange(tempC_dist, soilC_dist, 
                          ros_dist, ncol=2)
```

@@

```{r missing_table}
# MICE operations 
#
# Pattern of missing data
AnalysisData %>%
  select(-FireCode, -location, -block, -pasture,-patch,
         -year, -plot, -array, -LAI, -JD) %>%
    mice::md.pattern(plot = F) %>%
  pander::pander(caption="Table of missing values by variable.") 
```

@@

```{r imp, eval=FALSE}
# Calculate imputed datasets on scaled data
    imp_sc <- AnalysisData %>% 
            select(-LAI) %>%
            mutate_at(vars(AirTemp:tHa), ~as.numeric(scale(., center=F))) %>%
                      mutate(across(location:array, as.factor)) %>%
                       mice(m=5, seed = 23109, print=F)
# Reduce mids object to tibble for multivariate analysis    
    imp_raw <- 
      complete(imp, 'long') %>%
      as_tibble() %>%
      unite("TreeID", c(location, block, pasture,
                        year, plot, array), sep = ".") %>%
      select(-patch, -.id, -.imp, -FireCode) %>%
      pivot_longer(names_to = "response", 
                   values_to = "values", 
                   -TreeID) %>%
      group_by(TreeID, response) %>%
      summarize(value = median(values)) %>%
      ungroup() %>%
      pivot_wider(names_from = response, 
                  values_from = value) %>%
      separate(TreeID, c("location", "block", "pasture",
                         "year", "plot", "array")) %>% 
      mutate(across(location:array, as.factor))
```

@@

Mixed-effect regression models fit with `lme4::glmer` and constructed as follows:

@@

```{r glmer_mods, eval=FALSE}
# Mixed-effect regression models 
#
# Rate of spread
#   
  # Fit model 
    ros_imp <- with(imp_sc, suppressMessages(
                lme4::glmer(ros ~ RH + WindSpeed + 
                              FuelMoisture + tHa + 
                              (1|location/block/year/plot), 
                            family=Gamma(link = "log"), 
                            control=lme4::glmerControl(optimizer="bobyqa", 
                                                       optCtrl=list(maxfun=100000)) )) )
    ros_imp_pooled <- pool(ros_imp)$pooled  %>%
                          as_tibble()
    # Get terms
    ros_terms <- 
      full_join(
      summary(pool(ros_imp)) %>% 
        as_tibble() %>%
        rownames_to_column("row"),

      confint.mipo(pool(ros_imp)) %>%
        as_tibble() %>%
        rownames_to_column("row") )

#
# Maximum canopy temperature  
#
# Fit model 
canopy_imp <- with(imp_sc, suppressMessages(
              lme4::glmer(MaxC ~ RH + WindSpeed +  
                            FuelMoisture + tHa + 
                            (1|location/block/year/plot), 
                          family=Gamma(link = "log"), 
                          control=lme4::glmerControl(optimizer="bobyqa", 
                                                     optCtrl=list(maxfun=100000)) )) )

  canopy_imp_pooled <- pool(canopy_imp)$pooled  %>%
                      as_tibble()

  canopy_terms <- 
    full_join(
      summary(pool(canopy_imp)) %>% 
        as_tibble() %>%
        rownames_to_column("row"),

      confint.mipo(pool(canopy_imp)) %>%
        as_tibble() %>%
        rownames_to_column("row") )

#
# Maximum soil surface temperature  
#
# Fit model 
  soil_imp <- with(imp_sc, suppressMessages(
                  lme4::glmer(log(SoilMaxC+1) ~ RH + WindSpeed +  
                      FuelMoisture + tHa + 
                      (1|location/block/year/plot), 
                    family=Gamma(link = "log"), 
                    control=lme4::glmerControl(optimizer="bobyqa", 
                                               optCtrl=list(maxfun=100000)) )) )

soil_imp_pooled <- pool(soil_imp)$pooled  %>%
  as_tibble()

soil_terms <- 
  full_join(
    summary(pool(soil_imp)) %>% 
      as_tibble() %>%
      rownames_to_column("row"),

    confint.mipo(pool(soil_imp)) %>%
      as_tibble() %>%
      rownames_to_column("row") )

response_CIs <-   
  bind_rows(
    mutate(soil_terms, response = 'Temp. at surface'), 
    mutate(canopy_terms, response = 'Temp. above surface'),
    mutate(ros_terms, response = "Rate of spread") 
  ) %>%
  as_tibble() %>%
  rename(lwr = `2.5 %`, upr = `97.5 %`)
```

@@

### Rate of spread

```{r }
response_CIs %>%
  filter(response == "Rate of spread") %>%
  rename(t = statistic, 
         P = p.value) %>%
    select(term, t, P) %>%
  mutate(across(t:P, ~round(., 2))) %>%
  pander::pander("Summary statistics for Rate of spread.") 
```

### Aboveground temperatures

```{r }
response_CIs %>%
  filter(response == "Temp. above surface") %>%
  rename(t = statistic, 
         P = p.value) %>%
    select(term, t, P) %>%
  mutate(across(t:P, ~round(., 2))) %>%
  pander::pander("Summary statistics for flame temperature 15 cm above soil surface.") 
```

### Soil surface temperature

```{r }
response_CIs %>%
  filter(response == "Temp. at surface") %>%
  rename(t = statistic, 
         P = p.value) %>%
    select(term, t, P) %>%
  mutate(across(t:P, ~round(., 2))) %>%
  pander::pander("Summary statistics for soil surface temperature.") 
```

```{r CI_gg, fig.width=8, fig.height = 3.75}
# Plot confidence intervals
response_CIs %>%
  ggplot(aes(x = reorder(term, abs(estimate), max))) + theme_bw(16) + 
    geom_hline(yintercept = 0, color="black", linetype = 2) +
    geom_errorbar(aes(ymin = lwr, ymax = upr), 
                  size = 1.1, width = 0.2, 
                  color = "blue") +
    geom_point(aes(y = estimate), pch = 21, 
               stroke = 1.5, size = 4,
               color = "blue", fill="lightblue") +
    labs(x = '', 
         y = 'Regression coefficient with 95% CI') + 
    coord_flip(ylim=c(-1.5,2.4)) + 
    facet_wrap(~ response) + 
    theme(axis.text.y = element_text(color="black", size = 14), 
          strip.text = element_text(size = 14), 
          panel.grid.major.y = element_blank())
```
